\name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Model for a Precision Spectrograph

### 2020-05-08 paper/paper.tex
formatting tweaks; got modern to work
Incorporating MB Text Comments
Still missing some larger-ticket edits and figure changes

\documentclass[12pt, onecolumn]{aastex63}
\usepackage{xcolor}
\newcommand{\lz}[1]{\textcolor{orange}{#1}}

\usepackage[ruled,vlined]{algorithm2e}
\usepackage{graphicx}

% typesetting words
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\name}{\project{Excalibur}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\expres}{\project{\acronym{EXPRES}}}
\newcommand{\eprv}{\acronym{EPRV}}
\newcommand{\lfc}{\acronym{LFC}}
\newcommand{\code}[1]{\texttt{#1}}

@@

% margins and page setup, etc
\addtolength{\textheight}{1.00in}
\addtolength{\topmargin}{-0.50in}
\sloppy\sloppypar\raggedbottom\frenchspacing

\begin{document}
\title{\name:
A Non-Parametric, Hierarchical Wavelength-Calibration Model for a Precision Spectrograph}

@@

\begin{abstract}
\name\ is a hierarchical, non-parametric framework for wavelength calibrating precision spectrographs.  This method capitalizes on recent hardware improvements in highly-stabilized, extreme-precision radial-velocity spectrographs.  With stabilized spectrographs, the full optical system and detectors can only vary within a tiny range of configurations.  This allows us to use all calibration data to determine the entire space of possible calibrations for a stabilized spectrograph.  A wavelength solution then need only pinpoint where in this low-dimensional calibration space the spectrograph exists.  \name\ also takes advantage of the development of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density of points permits the freedom of a non-parametric wavelength solution that can match or adapt to any instrument or detector oddities, more so than any one polynomial.  We demonstrate the success of this method with data from the EXtreme PREcision Spectrograph, \expres, which includes a Menlo Systems laser frequency comb.  We find that \name\ returns more accurate wavelength predictions by more than a factor of 2 over classic, non-hierarchical, polynomial solutions.  Employing \name\ wavelengths on \expres\ RV measurements on HD 3651 reduced the RMS of a planet fit by \lz{blah}.
\end{abstract}

@@

\section{Introduction} 
Extremely precise, radial-velocity (\eprv) programs have been incredibly fruitful in finding and characterizing extra-solar planets \citep{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $10^5$, which correspond to line widths on the order of $3000\,\mps$.  Exoplanet science happens at the level of $3\,\mps$ precision, with current instruments hoping to reach $0.1\,\mps$ precision \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).    Some hardware--software systems, such as ESPRESSO, \expres\, and NEID were designed to deliver precision at this level \citep{pepe2013,  jurgenson2016, neid}.  Early data from \expres\ and ESPRESSO suggest that the hardware improvements do propagate through to sub-$\mps$ uncertainties on final, on-sky RV measurements \citep{blackman2020, petersburg2020, pepe2020}.

@@

Here we propose to simplify and improve calibration programs for \eprv\ hardware systems with two very simple but innovative ideas.  The first flows from the observation that calibration sources---which include arc lamps (in some wavelength ranges), etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines; almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.

@@

The second simple idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season, and only along a small number of axes in what you might call ``calibration space'', or the (very high dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.

In the context of probabilistic models, this structure is \emph{hierarchical}:  The calibration data are used not just to determine the wavelength solution, but also to determine the possible \emph{space} of wavelength solutions.  In the statistics literature this concept is often described as \emph{de-noising}:  Every calibration exposure contains information about every other exposure.  Thus every exposure can be improved (i.e. de-noised) with information from every other exposure.

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven model for the wavelength solution.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically exercised variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind the \eprv\ instruments and \eprv\ science cases, but we expect \name\ to have applications for other kinds of spectrographs in other contexts.  The motivating ideas behind this project are true for almost every astronomical spectrograph.

\section{Method} \label{sec:method}
\name\ is designed to turn a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  \name\ operates on the two core ideas that the wavelength solution should be given enormous flexibility, but that it lives in a very low-dimensional space, where the degrees of freedom are set by the limited kinematics of the stabilized spectrograph hardware.

\name\ assumes dense enough calibration line coverage with well-fit line centers to provide some constraint on an interpolated wavelength solution across an echelle order.  Additionally, \name\ assumes the instrument hardware is stabilized enough that the space of possible calibration states is low-dimensional.

@@

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} to be a 2D, 9\textsuperscript{th}-order polynomial where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
where the y-dependence is dropped as that dependence is carried by spectral order $m$.  The coefficients $c_{nij}$ are interpolated to time $t_n$ of exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains in $c_{nij}$ coefficients independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that than can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  When this calibration structure is used to deliver a wavelength solution, we will have to invert $x(\lambda,m,n)$ into $\lambda(x,m,n)$.

@@

\subsection{Dimensionality Reduction: De-Noising of Calibration Frames} \label{sec:denoising}
Within a given range of stability, \name\ will use calibration images to  1) determine the space in which the instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  For each calibration exposure, $n$, \name\ requires a full list of lines, $(\lambda,m)$ that are expected in each calibration exposure.  Each line is uniquely defined by a combination of echelle order, $m$, and ``true'' or theoretical wavelength, $\lambda$.

@@

Missing line center measurements are replaced with de-noised estimates.  This is done iteratively until the estimates of missing line centers change by less than 0.01\%.  This process can be repeated on line centers deemed as outliers by some metric, to account for lines that may have been mis-identified or mis-fit.  The principal components from the final iteration are used to define the spectrograph's calibration space while the associated magnitudes pinpoint where in that calibration space the spectrograph is located for each calibration exposure.

\begin{algorithm}
\SetAlgoLined
\KwData{line positions $x(\lambda,m,n)$ for each exposure $n$, with wavelengths $\lambda$ and echelle orders $m$}
\KwResult{Low-dimensional calibration space of spectrograph}
\While{change in missing or outlier line centers $>$ 0.01\%}
{
	$g_0(\lambda,m) = \overline{x(\lambda,m,n)}$\;
	find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x(\lambda,m,n)-g_0(\lambda,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(\lambda,m) = V$\;
	$x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)$ where $K=6$ for $x(\lambda,m,n)$ = \code{NaN}
	}
\caption{Dimension Reduction and De-Noising}
\end{algorithm}

\subsection{?Interpolating Calibration Frames?}
 \label{sec:interp_time}
The magnitude of each principal component can be interpolated with respect to housekeeping data, such as time, temperature, telescope position, etc. to determine different calibration states of the spectrograph.  The choice of what to interpolate against depends on the dominant contribution to variation in the instrument.

In the implementation of \name\ presented here, the magnitudes of the principal components are interpolated with respect to time.  New magnitudes $a_n'k$ values are found for the midpoint time of non-calibration exposures and used to construct the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a_{n;k}$ and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.

@@

@@

The results and discussion presented here are based off of 1227 LFC exposures taken between October 14 and December 18, 2020 on 29 unique nights.  LFC lines cover 40 echelle orders, which contain 19203 lines.  Though the results are primarily based on work with LFC data, there is some discussion of applications to arc lamps.  These results use 78 ThAr exposures that were taken on the same nights.  ThAr lines cover all 86 extracted orders of \expres, which includes 5295 lines.  For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, files with more than 60\% of expected lines missing were cut from the analysis.

 \name\ proceeds from a given list of echelle orders, line wavelengths, and pixel positions of each line for each exposure.  We, therefore, use line positions generated by the pre-existing \expres\ pipeline \citep{petersburg2020}.

 A ThAr wavelength solution is generated from each ThAr exposure using the IDL code \texttt{thid.pro}, developed by Jeff Valenti.  This code identifies ThAr lines by matching lines in an exposure against a line atlas.

 Flat-relative, optimally extracted LFC data is background-corrected using a univariate spline.  Each peak in an echelle order is then fit to a Gaussian.  The mean of the fitted Gaussian is taken to be the center of the line.  For each line, the ThAr wavelength solution is used to estimate the mode number of a line.  The precise wavelength is then calculated using
 \begin{equation}
 f_n = n \times  f_r + f_0
 \label{eq:lfc}
 \end{equation}
 where the repetition rate, $f_r$, is known from design of the LFC, and the offset frequency, $f_0$, has been determined experimentally.  the pre-existing \expres\ pipeline then generates wavelength solutions on a night-by-night basis as described in Section \ref{sec:method}.

@@

In order to satisfy the assumption of stability needed for \name, only exposures from only one ``instrumental epoch" of \expres\ are used.  \expres\ exposures are separated into different instrumental epochs that correspond to atypical hardware changes--changes that altered the position of the echellogram on the detector, the shape of the instrumental PSF, or the calibration sources.  Significant instrumental changes break the assumption of a stabilized instrument with only low-dimensional variations.  Each instrument epoch is therefore treated independently.

\section{Tests}\label{sec:tests}
We perform a series of tests to validate the performance of \name\ and benchmark \name -generated wavelengths against wavelengths generated by a classic, parametric method.  For all these tests, we generate wavelengths for known calibration lines and compare the predicted wavelength to the assigned wavelength of each line.  This inherently folds in errors in the measured line center of each calibration line, but the contribution to the residuals will be the same across all tests.

To assess the classic, polynomial-driven method of wavelength calibration, we take each LFC exposure and separate the lines into even and odd lines.  We then construct a wavelength solution using only the odd lines and use it to predict wavelengths for the even lines; i.e. a polynomial is fit to just the odd lines and then evaluated at the detector position of the even lines.  We then generate a wavelength solution using only the even lines and use it to predict wavelengths for the odd lines.

@@

Histograms of the residuals for each of the above described polynomial, interpolation, and denoising tests (respectively) are shown in Figure \ref{fig:testHists}.  Note that the spread in residuals is much smaller for both the denoising and interpolation tests relative to the results of the polynomial wavelength solutionl.

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and assigned wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The standard deviation of the residuals for each test is given in the top-right corner in each method's corresponding color.}
\label{fig:testHists}
\end{figure}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/lineResids2D.png}
%\includegraphics[width=\textwidth]{Figures/lineResids2D.pdf}
\caption{Residuals plotted with respect to detector position (as defined by echelle order and x-pixel) for various wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the true wavelength for each line, given in units of $/mps$.  Residuals to a classic, polynomial fit is shown in the left-most plot.  The center plot shows residuals to an interpolated wavelength solution.  The right-most plot shows residuals when a ThAr wavelength solution is interpolated onto LFC line positions.  The position of ThAr lines are over-plotted in black.}
\label{fig:resid2d}
\end{figure*}

@@

The residuals of the classic, polynomial wavelength solution is shown in the left-most plot of Figure \ref{fig:resid2d}.  There is a lot of vertical structure and some hints of a periodic, diagonal structure as well.

@@

\subsection{Value of K}
The value of $K$ represents the dimensionality of the calibration space within which the spectrograph lives.  In practice, it is the number of principal components, or basis vectors, used to reconstruct the de-noised line centers.  $K$ needs to be large enough so that all variability in the spectrograph is captured.  Too large, however, and the reconstruction will begin to incorporate noise, thereby defeating the purpose.

@@

In deciding a K value, we ran denoising tests, as described in Section \ref{sec:tests}, for K values spanning from 2 to 512.  One would expect the returned wavelengths to get increasingly more accurate with larger K until components that represent only noise are incorporated.  We found that the returned wavelengths have the lowest residuals with a K value of 32.  However, the improvement is slight with K values greater than six.  Comparisons of wavelengths returned by a K=6 model vs. a K=32 model show significant differences for less than 10, bluer lines.  A visual inspection of the principal components, such as the one displayed in Figure \ref{fig:pcLfc}, led us to believe most of the structure is being captured in the first six principal components.

\subsection{Interpolation of $a_{nk}$ with Respect to Time}
Figure \ref{fig:nightlyVariation} shows the amplitude of the first and second principal component with respect to time on the left.  Though there exists a complex overall shape to the amplitudes with respect to time, a clear linear trend exists within each night.  This is shown by the right plots in Figure  \ref{fig:nightlyVariation}.  As the beginning-of-night and end-of-night calibration sets always include LFC exposures, we use a simple linear interpolation to interpolate principal component amplitudes with respect to time.

@@

@@

\subsection{Interpolation of Wavelengths with Respect to Pixel}
Interpolation of wavelengths over pixel  is done order by order using a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator.  This interpolation incorporates the flexibility needed to model the changing dispersion of the spectrograph across an echelle order along with any detector defects while also enforcing monotonicity, which we know must be true across any one echelle order.

>>>>

\documentclass[modern]{aastex63}
\usepackage[ruled,vlined,norelsize]{algorithm2e}

% hogg dealing with his shit
\addtolength{\topmargin}{-0.30in}
\addtolength{\textheight}{0.70in}
\renewcommand{\twocolumngrid}{} % hahaha!
\setlength{\parindent}{1.1\baselineskip}
\sloppy\sloppypar\raggedbottom\frenchspacing
\shorttitle{excalibur: a wavelength-calibration model}
\shortauthors{zhao, hogg, bedell, fischer}

% typesetting words
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\name}{\project{Excalibur}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\expres}{\project{\acronym{EXPRES}}}
\newcommand{\espresso}{\project{\acronym{ESPRESSO}}}
\newcommand{\neid}{\project{\acronym{NEID}}}
\newcommand{\eprv}{\acronym{EPRV}}
\newcommand{\lfc}{\acronym{LFC}}
\newcommand{\code}[1]{\texttt{#1}}

\newcommand{\lz}[1]{\textcolor{orange}{#1}}
\newcommand{\mb}[1]{\textcolor{violet}{#1}}

@@

\begin{document}
\title{\name:
  A Non-Parametric, Hierarchical \\
  Wavelength-Calibration Model for a Precision Spectrograph}

@@

\begin{abstract}\noindent%
\name\ is a hierarchical, non-parametric framework for wavelength calibrating precision spectrographs.  This method capitalizes on recent hardware improvements in highly-stabilized, extreme-precision radial-velocity spectrographs.  With stabilized spectrographs, the full optical system and detectors can only vary within a tiny range of configurations.  This allows us to use all calibration data to determine the entire space of possible calibrations for a stabilized spectrograph.  A wavelength solution then need only pinpoint where in this low-dimensional calibration space the spectrograph exists.  \name\ also takes advantage of the development of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density of points permits the freedom of a non-parametric wavelength solution that can match or adapt to any instrument or detector oddities, more so than any one polynomial.  We demonstrate the success of this method with data from the EXtreme PREcision Spectrograph, \expres, which includes a Menlo Systems laser frequency comb.  We find that \name\ returns more accurate wavelength predictions by more than a factor of 2 over classic, non-hierarchical, polynomial solutions.  Employing \name\ wavelengths on \expres\ RV measurements on HD 3651 reduced the RMS of a planet fit by \lz{blah}.
\end{abstract}

@@

\section{Introduction} 
Extremely precise, radial-velocity (\eprv) programs have been incredibly fruitful in finding and characterizing extra-solar planets \citep{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $10^5$, which correspond to line widths on the order of $3000\,\mps$.  Exoplanet science happens at the level of $3\,\mps$ precision, with current instruments aiming to reach $0.1\,\mps$ precision \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).    Some hardware--software systems, such as \espresso, \expres, and \neid\ were designed to deliver precision at this level \citep{pepe2013,  jurgenson2016, neid}.  Early data from \expres\ and \espresso\ suggest that the hardware improvements do propagate through to sub-$\mps$ uncertainties on final, on-sky RV measurements \citep{blackman2020, petersburg2020, pepe2020}.

@@

Here we propose to simplify and improve calibration programs for \eprv\ hardware systems with two very simple but innovative ideas.  The first flows from the fact that calibration sources---which include arc lamps (in some wavelength ranges), etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines; almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.

@@

The second simple idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season.  \lz{This suggests that such stabilized spectrographs vary along only} a small number of axes in what you might call ``calibration space'', or the (very high dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.

In the context of probabilistic models, this structure is \emph{hierarchical}:  The calibration data are used not just to determine the wavelength solution at one moment, but also to determine the possible \emph{space} of wavelength solutions at all moments.  In the statistics literature this concept is often described as \emph{de-noising}:  Every calibration exposure contains information about every other exposure.  Thus every exposure can be improved (i.e. de-noised) with information from every other exposure.

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven model for the wavelength solution.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically determined variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases, but we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \mb{The motivating ideas behind this project are true for almost every astronomical spectrograph. (Is this true?)}

\section{Method} \label{sec:method}
The \name\ method is designed to turn a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  \name\ operates on the two core ideas that the wavelength solution should be given enormous flexibility, but that it lives in a very low-dimensional space, where the degrees of freedom are set by the limited kinematics of the stabilized spectrograph hardware.

\name\ assumes dense enough calibration line coverage with well-fit line centers to provide some constraint on an interpolated wavelength solution across an echelle order.  Additionally, \name\ assumes the instrument hardware is stabilized enough that the space of possible calibration states is low-dimensional.  We revisit and quantify these conditions in Section \ref{sec:discussion}.

@@

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} to be a 2D, 9\textsuperscript{th}-order polynomial where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here the y-dependence is dropped as that dependence is carried by spectral order $m$.  \lz{The line position information is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.}  The coefficients $c_{nij}$ are interpolated to time $t_n$ of exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that they can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  When this calibration structure is used to deliver a wavelength solution, we will have to invert $x(\lambda,m,n)$ into $\lambda(x,m,n)$.  \mb{(Explain why this inversion is necessary?)}

@@

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{Figures/methodCartoon.pdf}
\caption{}
\label{fig:testHists}
\end{figure*}

\subsection{Dimensionality Reduction: De-Noising of Calibration Frames} \label{sec:denoising}
Within a given range of stability, \name\ will use calibration images to  1) determine the space in which the instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  For each calibration exposure, $n$, \name\ requires a full list of lines, $(\lambda,m)$ that are expected in each calibration exposure.  Each line is uniquely defined by a combination of echelle order, $m$, and ``true'' or theoretical wavelength, $\lambda$.  \lz{There are many strategies for identifying calibration line positions and matching them to their assigned wavelengths; this problem is left out of scope for this work.}

@@

Missing line center measurements, which were previously marked by \code{NaN}s, are replaced with de-noised estimates.  This is done iteratively until the estimates of missing line centers change by less than 0.01\%.  This process can be repeated on line centers deemed as outliers by some metric, to account for lines that may have been mis-identified or mis-fit.  The principal components from the final iteration are used to define the spectrograph's calibration space while the associated magnitudes pinpoint where in that calibration space the spectrograph is located for each calibration exposure.

\begin{algorithm}
\SetAlgoLined
\KwData{line positions $x(\lambda,m,n)$ for each exposure $n$, with wavelengths $\lambda$ and echelle orders $m$}
\KwResult{Principle vectors of low-dimensional calibration space $g_k(\lambda,m)$, location of exposures in this space $a_{n,k}$}
\While{change in missing or outlier line centers $>$ 0.01\%}
{
	$g_0(\lambda,m) = \overline{x(\lambda,m,n)}$\;
	find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x(\lambda,m,n)-g_0(\lambda,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(\lambda,m) = V$\;
	$x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)$ where $K=6$ for $x(\lambda,m,n)$ = \code{NaN}
	}
\caption{Dimension Reduction and De-Noising}
\end{algorithm}

\subsection{?Interpolating Calibration Position?}
 \label{sec:interp_time}
The magnitude of each principal component can be interpolated with respect to housekeeping data, such as time, temperature, telescope position, etc. to determine different calibration states of the spectrograph.  The choice of what to interpolate against depends on the dominant contribution to variation in the instrument.

In the implementation of \name\ presented here, the magnitudes of the principal components are interpolated with respect to time.  New magnitudes $a_n'k$ are found for the midpoint time of non-calibration exposures and used to construct the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a_{n;k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.  \lz{The best interpolation scheme for \expres\' instrument drift is discussed below in Section \ref{sec:choice_avt}, but may vary for different instruments.}

@@

We found that a cubic-spline interpolation that enforces monotonicicity, such as a Piecewise Cubic Hermite Interpolatin Polynoimal (PCHIP) interpolator, works well for interpolating wavelengths onto pixels.  A cubic spline allows for more flexibility than a continuous function, while the enforced monotonicity prevents large deviations that may befall a cubic spline.  Choice in interpolation scheme and other tests are further discussed in Section \ref{sec:choice_wvp}.

@@

LFC lines cover 40 echelle orders, which contain 19203 lines.  Though the results are primarily based on work with LFC data, there is some discussion of applications to arc lamps.   ThAr lines cover all 86 extracted orders of \expres, which includes 5295 lines.

For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, files with more than 60\% of expected lines missing were cut from the analysis.

 \name\ requests as input a list of echelle orders, line wavelengths, and pixel positions of each line for each exposure.  We, therefore, use line positions generated by the pre-existing \expres\ pipeline from the following procedure \citep{petersburg2020}.

 A ThAr wavelength solution is generated from each ThAr exposure using the IDL code \texttt{thid.pro}, developed by Jeff Valenti.  This code identifies ThAr lines by matching lines in an exposure against a line atlas.  Lines are matched manually when there is a significant change in the instrument, but is otherwise done using forward-modeling.  Once each line's position is identified and matched to a wavelength from the line atlas, a sixth-order, 2D polynomial is fit over pixel location $x$, echelle order $m$, and scaled wavelength $m\lambda$ (in order to distinguish lines that may appear in more than one order).

 Flat-relative, optimally extracted LFC data is background-corrected using a univariate spline.  Each peak in an echelle order is then fit to a Gaussian.  The mean of the fitted Gaussian is taken to be the center of the line.  For each line, the ThAr wavelength solution is used to estimate the mode number of a line.  The precise wavelength is then calculated using
 \begin{equation}
 f_n = n \times  f_r + f_0
 \label{eq:lfc}
 \end{equation}
 where the repetition rate, $f_r$, is known from design of the LFC, and the offset frequency, $f_0$, has been determined experimentally.

@@

In order to satisfy the assumption of stability needed for \name, only exposures from only one ``instrumental epoch" of \expres\ are used.   \expres\ exposures are separated into different instrumental epochs that correspond to atypical hardware changes--changes that altered the position of the echellogram on the detector, the shape of the instrumental PSF, or the calibration sources.  Significant instrumental changes break the assumption of a stabilized instrument with only low-dimensional variations.  Each instrument epoch is therefore treated independently.  In the results presented here, we use 1227 LFC exposures and 78 ThAr exposures taken between October 14 and December 18, 2020 on 29 unique nights.

\section{Tests}\label{sec:tests}
We perform a series of tests to validate the performance of \name\ and benchmark \name -generated wavelengths against wavelengths generated by a classic, parametric method.  For all these tests, we leave out a subset of calibration lines with known wavelengths as a ``validation'' sample, generate wavelengths for these lines, and compare the predicted wavelength to the assigned wavelength of each line.  This inherently folds in errors in the measured line center of each calibration line, but this contribution to the residuals will be the same across all tests.

To assess the classic, polynomial-driven method of wavelength calibration, we take each LFC exposure and separate the lines into even and odd lines.  We then construct a wavelength solution using only the odd lines and use it to predict wavelengths for the even lines; i.e. a polynomial is fit to just the odd lines and then evaluated at the detector position of the even lines (see Equation \ref{eq:poly_wsol}).  We then generate a wavelength solution using only the even lines and use it to predict wavelengths for the odd lines.

@@

Histograms of the residuals for each of the above described polynomial, interpolation, and denoising tests (respectively) are shown in Figure \ref{fig:testHists}.  Note that the spread in residuals is much smaller for both the denoising and interpolation tests relative to the results of the polynomial wavelength solution.

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{Figures/all_results.pdf}
\caption{\lz{(change to be b/w safe)} Difference in predicted and assigned wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The standard deviation of the residuals for each test is given in the top-right corner in each method's corresponding color.}
\label{fig:testHists}
\end{figure}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/lineResids2D.png}
%\includegraphics[width=\textwidth]{Figures/lineResids2D.pdf}
\caption{Residuals for a single image plotted with respect to detector position (as defined by echelle order and x-pixel) for various wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the true wavelength for each line, given in units of $/mps$.  Residuals to a classic, polynomial fit is shown in the left-most plot.  The center plot shows residuals to an interpolated wavelength solution.  The right-most plot shows residuals when a ThAr wavelength solution is interpolated onto LFC line positions.  The position of ThAr lines are over-plotted in black.}
\label{fig:resid2d}
\end{figure*}

@@

The residuals of the classic, polynomial wavelength solution is shown in the left-most plot of Figure \ref{fig:resid2d}.  There is a lot of vertical structure and some hints of a periodic, diagonal structure as well.  \lz{(Possibly from detector effects?  We're double checking this now and thinking about approaching STA about it.  Debra was considering publishing the results from that discussion in this paper if it seems relevant.)}.

@@

\subsection{Value of K}
\label{sec:choice_k}
The value of $K$ represents the dimensionality of the calibration space within which the spectrograph lives.  In practice, it is the number of principal components, or basis vectors, used to reconstruct the de-noised line centers.  $K$ needs to be large enough so that all variability in the spectrograph is captured.  Too large, however, and the reconstruction will begin to incorporate noise, thereby defeating the purpose.

@@

In deciding a K value, we ran denoising tests, as described in Section \ref{sec:tests}, for K values spanning from 2 to 512.  One would expect the returned wavelengths to get increasingly more accurate with larger K until components that represent only noise are incorporated.  We found that the returned wavelengths have the lowest residuals with a K value of 32 (see Figure \ref{fig:kvals}).  However, the improvement is slight with K values greater than six.  Comparisons of wavelengths returned by a K=6 model vs. a K=32 model show significant differences for less than 10, bluer lines.  A visual inspection of the principal components, such as the one displayed in Figure \ref{fig:pcLfc}, led us to believe most of the structure is being captured in the first six principal components.

\begin{figure}[t]
\centering
\includegraphics[width=.1\textwidth]{Figures/pcsLfc6.png}
\caption{Spread in residuals as a function of $K$.}
\label{fig:kvals}
\end{figure}

\subsection{Interpolation of $a_{nk}$ with Respect to Time}
\label{sec:choice_avt}
Figure \ref{fig:nightlyVariation} shows the amplitude of the first and second principal component with respect to time on the left.  Though there exists a complex overall shape to the amplitudes with respect to time, a clear linear trend exists within each night.  This is shown by the right plots in Figure  \ref{fig:nightlyVariation}.  As the beginning-of-night and end-of-night calibration sets always include LFC exposures, we use a simple linear interpolation to interpolate principal component amplitudes with respect to time.

@@

\mb{(Can we quantify or illustrate how much change is too much change, for use in discussion of other instruments?)}

@@

\lz{Add empirical determination of how often EXPRES needs to redo calibration exposures?}

\subsection{Interpolation of Wavelengths with Respect to Pixel}
\label{sec:choice_wvp}
Interpolation of wavelengths over pixel  is done order by order using a Piecewise Cubic Hermite Interpolating Polynomial (PCHIP) interpolator.  This interpolation incorporates the flexibility needed to model the changing dispersion of the spectrograph across an echelle order along with any detector defects while also enforcing monotonicity, which we know must be true across any one echelle order.
